\prefacesection{Acknowledgments}

Without a doubt, I would not be here at Stanford without my advisor Mykel Kochenderfer.
Going back to 2012 when we first met at MIT Lincoln Laboratory, you captured an excitement in me that has allowed me to flourish as a researcher.
Your style of advising and constant care for the well-being of your students is incredibly inspiring.
Whether I am having a stressful week, productive week, or thrilled to show you \textit{``one more thing,''} you have given me the space and mentorship to be a completely open and honest student.
I would also like to recognize my CS co-advisor, Clark Barrett, my unofficial Earth and Planetary Sciences advisor, Jef Caers, and my defense committee, Mert Pilanci and Tapan Mukerji, for their encouragement, thoughtfulness, and constant reminder of what my primary goals as an objective researcher are.
I am also grateful for Dorsa Sadigh for being on my master's thesis committee and providing valuable feedback that led me to be a successful PhD student.

This work is an accumulation of years of discussion with my peers.
I am indebted to the following brilliant minds I have had the pleasure of collaborating with:
Anthony Corso, Ritchie Lee, Arec Jamgochian, Sydney Katz, Markus Zechner, Johannes Fischer, David Zhen Yin, Kyla Guru, and Mariia Kozlova.
I am incredibly lucky to be a part of the \textit{Stanford Intelligent Systems Laboratory} (SISL) and want to thank the following for their useful discussions and ideas:
Dylan Asmar, Harrison Delecki, Anka Reuel, John Mern, Bernard Lange, Josh Ott, Chelsea Sidrane, Liam Kruse, Marc Schlichting, Romeo Valentine, Grace Kim, Amelia Hardy, Houjun Liu, Lisa Einstein, Duncan Eddy, Kiana Jafari, Mansur Arief, Edward Balaban, Tomar Arnon, Jayesh Gupta, Mark Koren, Shushman Choudhury, Jean-Raymond Betterton, Ellen Xu, Alex Kufos, and Zach Sunberg.
I also want to shoutout Ashwim Kanhere for his continual support and willingness to review my papers.

I have also had the pleasure to work with James G. O'Brien as an undergraduate researcher while at Wentworth and I'm grateful for the inspiration and trust he provided me.
I owe an enormous debt to everyone I have interacted with while at MIT Lincoln Laboratory over the years---they each played a role in shaping me into a better scientist, engineer, analyst, and friend.
This includes Ted Londner, Michael Owen, Wes Olson, Rod Cole, Gregg Shoults, Ian Jessen, Adam Panken, Tomas Elder, Cindy McLain, Luis Alvarez, Tan Trinh, Robert Klaus, Justin MacKay, Jared Wikle, Larry Capuder, Emilie Cowen, Randy Wicken, Tom Teller, Matt Edwards, Adam Gjersvik, Sam Wu, Anshu Das, Jack Lepird, Charles Leeper, and Dan Griffith.
Particular appreciation goes to Jeff Bezanson for co-creating Julia and showing me its divinity.
Also a special thanks to Neal Suchy from the Federal Aviation Administration for his admiration and his faith in my work.

Research sponsorship provided most of the funding for this dissertation.
I would like to thank Maxime Gariel and Arthur Dubois for their willingness to let me explore ideas relating to my thesis while working at Xwing, and Rob Timpe and Alex Bridi for the development of the runway detection system that I study in this work.
I would like to thank OMV and, in particular, Torsten Clemens for their research support and always being excited about the directions we take our work. 
I am grateful for the support from Ideon to investigate an exciting application of POMDPs to muon-based intrusion discovery, and I am particularly grateful for Patrick Belliveau, Nigel Phillips, Doug Schouten, and Max Howarth.
Stanford research centers have also been incredibly supportive during my PhD, and I would like to thank the Mineral-X Initiative, the Stanford Center for AI Safety, the Stanford Institute for Human-Centered AI (HAI), and the Stanford Center for Earth Resources Forecasting (SCERF) as part of the Doerr School of Sustainability.

To my friends and family, I am extremely grateful that you have given me the opportunity to pursue a PhD.
Mom and Dad, thank you for always supporting every decision I make in this uncertain world and thank you Travis, Emily, and Jake for always telling me how proud you are to call me your brother.
To my close friends back in Rockport, MA, you all have seen me throughout each of my phases of life and continue to remind me to be a better person.

Most importantly, I need to thank my wife, Eva, for being my foundation.
A PhD is never something you do alone, and no one knows best like the spouse of a student.
Eva, you endured this stressful adventure with me, and I cannot express how thankful I am that you have given me the space to be selfish in my studies.
I am excited to share the rest of my life with you and our dog, Tzvi, and I look forward to the next generation of Mosses.
